|

### 2023-02-28 paper_comparison/paper_comparison/paper_comparison.Rmd
update power desc

# Caculation of p-value and power

## $p$-value

In hypothesis testing, a $p$-value is defined as the probability of observing test results as least as extreme as the observed result given $H_0$ is true. Within the context of visual inference, by involving $k$ independent observers, the $p$-value can be interpreted as the probability of having as many or more subjects detect the data plot than the observed result.

@@

<!-- As pointed out by @vanderplas2021statistical, this basic binomial model doesn't take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup. And it is inapplicable to visual test where subjects are asked to select one or more "most different" plots from the lineup. @vanderplas2021statistical summarises three common scenarios in visual inference: (1) $K$ different lineups are shown to $K$ subjects, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ subjects, and (3) the same lineup is shown to $K$ subjects. Out of these three scenarios, Scenario 3 is the most common in previous studies as it puts the least constraints on the experiment design. For Scenario 3, @vanderplas2021statistical models the probability of a plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim Dirichlet(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case subject $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots subject $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual p-value is given as -->

<!-- <!-- TODO:Clarify this is only for integer c_i --> -->

<!-- ```{=tex} -->
<!-- \begin{equation} \label{eq:pvalue-beta-binomial} -->
<!-- P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+ -->
<!-- \end{equation} -->
<!-- ``` -->

<!-- \noindent where $B(.)$ is the beta function defined as -->

<!-- ```{=tex} -->
<!-- \begin{equation} \label{eq:betafunction} -->
<!-- B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.  -->
<!-- \end{equation} -->
<!-- ``` -->

<!-- Note that Equation \ref{eq:pvalue-beta-binomial} given in @vanderplas2021statistical only works with non-negative integer $c_i$. We extend the equation to non-negative real number $c_i$ by applying a linear approximation -->

<!-- ```{=tex} -->
<!-- \begin{equation} \label{eq:pvalue-beta-binomial-approx} -->
<!-- P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+, -->
<!-- \end{equation} -->
<!-- ``` -->

<!-- where $P(C \geq \lceil c_i \rceil)$ is calculated using Equation \ref{eq:pvalue-beta-binomial} and $P(C = \lfloor c_i \rfloor)$ is calculated by -->

<!-- \begin{equation} \label{eq:pmf-beta-binomial} -->
<!-- P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+. -->
<!-- \end{equation} -->

<!-- Besides, the parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and hence needs to be estimated from the survey data. For low values of $\alpha$, only a few plots are attractive to the observers and tend to be selected. For higher values of $\alpha$, the distribution of the probability of each plot being selected is more even. @vanderplas2021statistical defines that a plot is $c$-interesting if $c$ or more participants select the plot as the most different. Given the definition, The expected number of plots selected at least $c$ times, $E[Z_c]$, is calculated as -->

<!-- ```{=tex} -->
<!-- \begin{equation} \label{eq:c-interesting-expectation} -->
<!-- E[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation} -->
<!-- ``` -->

<!-- With Equation \ref{eq:c-interesting-expectation}, $\alpha$ can be estimated using maximum likelihood estimation. But for precise estimate of $\alpha$, additional responses to Rorschach lineups, which is a type of lineup that consists of plots constructed from the same null data generating mechanism, are required. -->

@@

<!-- The power of a model misspecification test is the probability that the null hypothesis is rejected given the regression model is misspecified. It is an important indicator when one is concerned about whether model assumptions have been violated. Although in practice, one might be more interested in knowing how much the residuals deviate from the model assumptions, and whether this deviation is of practical significance. -->

<!-- As discussed in @majumder_validation_2013, the power of a visual test may depend on the ability of the particular subject, as the skill of the individual may affect the number of observers who identify the data plot from the lineup. Previously, it is addressed by modelling the probability of a subject $j$ correctly picking the data plot from a lineup $l$ using a mixed-effect logistic regression, with subjects treated as random effects [@majumder_validation_2013]. However, in the multiple selections scenario, having this probability is not sufficient to determine the power of a visual test because it does not provide information about the number of selections made by the subject for the calculation of the p-value.  -->

<!-- Instead, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. The assumption is to simplify the model structure, thereby obviate costly large-scale experiments to estimate complex covariance matrices. The model is a logistic regression formulated as  -->

<!-- ```{=tex} -->
<!-- \begin{equation} \label{eq:logistic-regression-1-1} -->
<!-- Pr(\text{reject}~H_0|H_1,E) = \Lambda\left(log\left(\frac{0.05}{0.95}\right) + \beta_1 E\right), -->
<!-- \end{equation} -->
<!-- ``` -->
<!-- \noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$. The effect size $E$ is the only regressor and the intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$. -->

<!-- Effect size $E$ is formulated as the conditional expectation of the sample-based effect size $E_{sample}$ derived from the Kullback-Leibler divergence (see Appendix for the details) given by -->

<!-- \footnotesize -->

<!-- \begin{gather}  \label{eq:effect-size} -->
<!-- E = E[E_{sample}|M],\\ -->
<!-- E_{sample} = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a\sigma^2)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a\sigma^2)) + \boldsymbol{\mu}_z^{T}(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}\boldsymbol{\mu}_z\right),\\ -->
<!-- \boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a,\\ -->
<!-- \boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a^T\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a^T, -->
<!-- \end{gather} -->

<!-- \normalsize -->

<!-- \noindent where $M$ is a given model that generates the data, $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,  $\boldsymbol{X}_a = (\boldsymbol{1},  \boldsymbol{X})$ is the matrix of regressors including the intercept used in the regression equation, $\sigma^2\boldsymbol{I}$ is the assumed variance of the error term when $H_0$ is true, $\boldsymbol{V}$ is the actual variance of the error term, and $\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of the residuals with $\boldsymbol{Z}$ be any higher order terms of $\boldsymbol{X}$ leave out by the regression equation and $\boldsymbol{\beta}_z$ be the corresponding coefficients.  -->

<!-- A conditional expectation is taken because $E$ is designed to be invariant to different realizations of $\boldsymbol{X}$. However, solving this conditional expectation analytically is very difficult as $\boldsymbol{X}$ can follow arbitrary distributions. Therefore, we make an assumption that $\boldsymbol{X}$ follows a particular multivariate normal distribution -->

<!-- We make an assumption that all the realizations of $\boldsymbol{X}$ are generated from a multivariate normal distribution $N(\boldsymbol{0}, 0.3^2\boldsymbol{I})$ and then rescaled by the min-max normalization. The estimate of $E$ is obtained by taking the average of the $E_{sample}$ calculated for samples drawn from the given model $M$ repeatedly. We use the normal distribution as a general choice to approximate the conditional effect size. -->

<!-- To study various factors contributing to the power of the visual test, the same logistic regression model is fit on different subsets of the collated data grouped by levels of factors. These include the distribution of the regressor, type of the simulation model and the shape of the residual departures. -->

@@

>>>>

# Estimation of p-value and power

## p-value

In hypothesis testing, a $p$-value is defined as the probability of observing test results as least as extreme as the observed result given $H_0$ is true. Conventional hypothesis tests usually have an existing method to derive or compute $p$-value based on the null distribution. What we need to discuss in the following is the method to estimate $p$-value for a visual test.

Within the context of visual inference, by involving $k$ independent observers, the visual $p$-value can be interpreted as the probability of having as many or more subjects detect the data plot than the observed result.

@@

As pointed out by @vanderplas2021statistical, this basic binomial model doesn't take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup. And it is inapplicable to visual test where subjects are asked to select one or more "most different" plots from the lineup. @vanderplas2021statistical summarises three common scenarios in visual inference: (1) $K$ different lineups are shown to $K$ subjects, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ subjects, and (3) the same lineup is shown to $K$ subjects. Out of these three scenarios, Scenario 3 is the most common in previous studies as it puts the least constraints on the experiment design. For Scenario 3, @vanderplas2021statistical models the probability of a plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim Dirichlet(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case subject $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots subject $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual p-value is given as

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}
```

\noindent where $B(.)$ is the beta function defined as

```{=tex}
\begin{equation} \label{eq:betafunction}
B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.
\end{equation}
```

Note that Equation \ref{eq:pvalue-beta-binomial} given in @vanderplas2021statistical only works with non-negative integer $c_i$. We extend the equation to non-negative real number $c_i$ by applying a linear approximation

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}
```

where $P(C \geq \lceil c_i \rceil)$ is calculated using Equation \ref{eq:pvalue-beta-binomial} and $P(C = \lfloor c_i \rfloor)$ is calculated by

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and hence needs to be estimated from the survey data. For low values of $\alpha$, only a few plots are attractive to the observers and tend to be selected. For higher values of $\alpha$, the distribution of the probability of each plot being selected is more even. @vanderplas2021statistical defines that a plot is $c$-interesting if $c$ or more participants select the plot as the most different. Given the definition, The expected number of plots selected at least $c$ times, $E[Z_c]$, is calculated as

```{=tex}
\begin{equation} \label{eq:c-interesting-expectation}
E[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}
```

With Equation \ref{eq:c-interesting-expectation}, $\alpha$ can be estimated using maximum likelihood estimation. But for precise estimate of $\alpha$, additional human responses to Rorschach lineups, which is a type of lineup that consists of plots constructed from the same null data generating mechanism, are required.

@@

The power of a model misspecification test is the probability that the null hypothesis is rejected given the regression model is misspecified in a specific way. It is an important indicator
when one is concerned about whether model assumptions have been violated. Although in practice, one might be more interested in knowing how much the residuals deviate from the model assumptions, and whether this deviation is of practical significance.

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the effect size is provided in Section \ref{effect-size} after the introduction of the simulation model used in our human subject experiment. The theoretical power of a test is sometimes not a trival solution. For a given regression model under a specific effect size, if the data generating process is also known, the power of a conventional test can be estimated by counting the number of times the test rejects the null hypothesis in a repeated sampling exercise
and divided by the number of times simulated from the known data generating process. This is the method we used to estimate the power of conventional tests in our experiment.

The power of a visual test on the other hand, may additionally depend on the ability of the particular subject, as the skill of the individual may affect the number of observers who identify the data plot from the lineup [@majumder_validation_2013]. To address this issue, @majumder_validation_2013 models the probability of a subject $j$ correctly picking the data plot from a lineup $l$ using a mixed-effect logistic regression, with subjects treated as random effect. Then, the estimated power of a visual test evaluated by a single subject is the predicted value obtained from the mix-effect model.

However, this mix-effect model does not work with scenario where subjects are asked to select one or more "most different" plots. In this scenario, having the probability of a subject $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the subject for the calculation of the p-value (See Equation \ref{eq:pvalue-beta-binomial-approx}).

Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption is not necessary true, but it helps simplifying the model structure, thereby obviate a costly large-scale experiment to estimate complex covariance matrices. There is a discussion provided in the Appendix about a potential solution to model individual effect in the multiple selections scenario.

(should we just link the power together for conventional test?)

We use the same logistic regression model to fit the power of conventional tests and visual tests formulated as

```{=tex}
\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda\left(log\left(\frac{0.05}{0.95}\right) + \beta_1 E\right),
\end{equation}
```
\noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$. The effect size $E$ is the only regressor and the intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$.

To study various factors contributing to the power of test, the same logistic regression model is fit on different subsets of the collated data grouped by levels of factors. These include the distribution of the regressor, type of the simulation model and the shape of the residual departures.

@@

Effect size $E$ is formulated as the conditional expectation of the sample-based effect size $E_{sample}$ derived from the Kullback-Leibler divergence (see Appendix for the details) given by

\footnotesize

\begin{gather}  \label{eq:effect-size}
E = E[E_{sample}|M],\\
E_{sample} = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a\sigma^2)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a\sigma^2)) + \boldsymbol{\mu}_z^{T}(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}\boldsymbol{\mu}_z\right),\\
\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a,\\
\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a^T\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a^T,
\end{gather}

\normalsize

\noindent where $M$ is a given model that generates the data, $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,  $\boldsymbol{X}_a = (\boldsymbol{1},  \boldsymbol{X})$ is the matrix of regressors including the intercept used in the regression equation, $\sigma^2\boldsymbol{I}$ is the assumed variance of the error term when $H_0$ is true, $\boldsymbol{V}$ is the actual variance of the error term, and $\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of the residuals with $\boldsymbol{Z}$ be any higher order terms of $\boldsymbol{X}$ leave out by the regression equation and $\boldsymbol{\beta}_z$ be the corresponding coefficients.

A conditional expectation is taken because $E$ is designed to be invariant to different realizations of $\boldsymbol{X}$. However, solving this conditional expectation analytically is very difficult as $\boldsymbol{X}$ can follow arbitrary distributions. Therefore, we make an assumption that $\boldsymbol{X}$ follows a particular multivariate normal distribution

We make an assumption that all the realizations of $\boldsymbol{X}$ are generated from a multivariate normal distribution $N(\boldsymbol{0}, 0.3^2\boldsymbol{I})$ and then rescaled by the min-max normalization. The estimate of $E$ is obtained by taking the average of the $E_{sample}$ calculated for samples drawn from the given model $M$ repeatedly. We use the normal distribution as a general choice to approximate the conditional effect size.
